---
author: "EmmaLi Tsai, w/ help from Claude"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: lumen
params:
  # Column name in EJScreen (e.g. OZONE, PM25, DSLPM)
  ind: "OZONE"
  # Pipeline slug matching the indicator config file name (e.g. o3, pm25)
  ind_slug: "o3"
  # Pipeline version (e.g. 1.2022, 1.2021)
  ind_version: "1.2022"
  # EJScreen vintage year — used to name the baseline archive CSV
  ejscreen_year: "2024"
  # "local" or "remote" — controls whether paths resolve to disk or S3
  environment: "remote"
title: "EJScreen Comparison Report: `r params$ind`"
---

```{r cli-examples, include=FALSE}
# TODO - these CLI examples are all a work in progress! 
# TODO - use venvs
# TODO - resolve the remote vs local argument so it actually pulls the correct files


## CLI Examples
# Rscript -e "rmarkdown::render('./scripts/utilities/validation/compare_results.Rmd', params = list(ind = 'OZONE', ind_slug = 'o3', ind_version = '1.2022', ejscreen_year = '2024', environment = 'remote'))"

# Rscript -e "rmarkdown::render('./scripts/utilities/validation/compare_results.Rmd', params = list(ind = 'PM25',  ind_slug = 'pm25', ind_version = '1.2022', ejscreen_year = '2024', environment = 'remote'))"
```

```{r setup, include=FALSE}
library(tidyverse)
library(aws.s3)
library(tigris)
library(sf)
library(mapview)
library(sysfonts)
library(showtext)
library(purrr)
library(leaflet)
library(knitr)
library(kableExtra)
library(ggpmisc)
library(reticulate)
library(RColorBrewer)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

options(tigris_use_cache = TRUE)
options(scipen = 9999)
options(rmarkdown.pandoc.args = c("+RTS", "-K2g", "-RTS"))
mapviewOptions(basemaps = c("Esri.WorldGrayCanvas"))

VAR_NAME <- params$ind
```

```{r themes, include=FALSE}
# themes and color palettes
sysfonts::font_add_google("Lato")
showtext::showtext_auto()

epic_chart_theme <- theme_minimal() +
  theme(legend.position  = "right",
        text             = element_text(size = 13, family = "Lato"),
        legend.text      = element_text(size = 10, family = "Lato"),
        legend.title     = element_text(size = 11, family = "Lato"),
        axis.text.x      = element_text(margin = margin(t = 10)),
        axis.title.x     = element_text(margin = margin(t = 10)),
        axis.text.y      = element_text(margin = margin(r = 10)),
        axis.title.y     = element_text(margin = margin(r = 10)))

cat_palette_pastel <- colorRampPalette(
  c("#526489", "#527CAF", "#B077B2", "#E4BE7C", "#b15712", "#82AB6E")
)
cont_palette <- colorRampPalette(c("#172f60", "#4ea324"))
```

```{r helpers, include=FALSE}
# function for computing summary & comparison stats for columns a and b
compute_stats <- function(df, col_b, col_a) {
  valid     <- df %>%
    select(all_of(c(col_a, col_b))) %>%
    mutate(across(everything(), as.numeric))
  diffs     <- valid[[col_a]] - valid[[col_b]]
  pos_diffs <- diffs[!is.na(diffs) & diffs > 0]
  neg_diffs <- diffs[!is.na(diffs) & diffs < 0]

  tibble(
    # first, get basic min/median/mean/max for a and b
    min_a = min(valid[[col_a]], na.rm = T),
    median_a = median(valid[[col_a]], na.rm = T), 
    mean_a = mean(valid[[col_a]], na.rm = T),
    max_a = max(valid[[col_a]], na.rm = T),
    
    min_b = min(valid[[col_b]], na.rm = T),
    median_b = median(valid[[col_b]], na.rm = T), 
    mean_b = mean(valid[[col_b]], na.rm = T),
    max_b = max(valid[[col_b]], na.rm = T),
    
    # how many rows are missing information? 
    num_NA_var_a       = sum(is.na(valid[[col_a]])),
    num_NA_var_b       = sum(is.na(valid[[col_b]])),
    
    # comparison stats - looking at the diff between column A  - column B 
    mean_abs_diff     = mean(abs(diffs), na.rm = TRUE),
    median_abs_diff   = median(abs(diffs), na.rm = TRUE),
    rmse              = sqrt(mean(diffs^2, na.rm = TRUE)),
    pearson_r         = cor(valid[[col_a]], valid[[col_b]], 
                            use = "na.or.complete"),
    r_sq              = pearson_r^2,
    greatest_pos_diff = if (length(pos_diffs)) max(pos_diffs) else NA_real_,
    smallest_neg_diff = if (length(neg_diffs)) min(neg_diffs) else NA_real_
  )
}


# ---------------------------------------------------------------------------
# Resolve all file paths via the shared Python resolve_path module so that
# no paths are hard-coded in this script.
#
# Mirrors the pattern used in shared_preprocess.R:
#   py_build_manifest <- import_from_path("build_manifest", path = shared_path)
#
# For path_old (original EJScreen baseline archive):
#   - Local:  {repo_root}/outputs/EJSCREEN_{year}_BG_with_AS_CNMI_GU_VI.csv
#   - Remote: not yet in shared_config.json — add an "ejscreen_archive" asset
#             entry when the file is uploaded to S3, then switch to
#             rp$get_shared_asset_path(...) here.
#
# For path_new (pipeline ejscreen export):
#   - Derived from pipeline root + known folder convention:
#     ejscreen_acs_{ejscreen_year}/{ind_slug}_v{ind_version}/ejscreen_export.csv
# ---------------------------------------------------------------------------

# Walk up from this Rmd's directory until we find the .git anchor
find_repo_root <- function(start = NULL) {
  path <- normalizePath(if (is.null(start)) getwd() else start)
  while (!file.exists(file.path(path, ".git"))) {
    parent <- dirname(path)
    if (parent == path) stop("Could not locate repository root (.git)")
    path <- parent
  }
  path
}

# knitr sets the working directory to the Rmd's folder at render time
repo_root   <- find_repo_root(knitr::opts_knit$get("output.dir") %||% getwd())
scripts_dir <- file.path(repo_root, "scripts")
shared_dir  <- file.path(scripts_dir, "shared")

# Import resolve_path (same approach as shared_preprocess.R → import_from_path)
rp <- reticulate::import_from_path("resolve_path", path = shared_dir)

pipeline_root <- rp$get_pipeline_root(params$environment)

# New pipeline output — ejscreen export CSV produced by the merge step
path_new <- paste0(
  pipeline_root,
  paste0(
    "ejscreen_acs_", params$ejscreen_year, "/",
    params$ind_slug, "_v", params$ind_version, "/ejscreen_export.csv"
  )
)

# Original EJScreen baseline archive - defaults to always 
# pulling from AWS
path_old <- "s3://pedp-data-preserved/ejscreen-data-processing/pipeline/shared/ejscreen/EJSCREEN_2024_BG_with_AS_CNMI_GU_VI.csv"

message("path_old -> ", path_old)
message("path_new -> ", path_new)
```

```{r read-data, include=FALSE, cache=TRUE, message=FALSE}
# ---------------------------------------------------------------------------
# Memory-efficient reads: peek at headers first, then load only the columns
# that match VAR_NAME (+ ID and ST_ABBREV). For a 200-column EJScreen CSV
# this cuts the in-memory footprint by ~20x.
# ---------------------------------------------------------------------------

# --- OLD (defaults to always remote) ---
ejscreen_old <- aws.s3::s3read_using(read.csv, 
                                     object = path_old)

col_names_old <- names(ejscreen_old)
keep_old <- col_names_old[grepl(VAR_NAME, col_names_old, ignore.case = TRUE)]
ejscreen_old <- ejscreen_old %>%
  select(c("ID", "ST_ABBREV", all_of(keep_old))) %>%
  mutate(ID = as.character(ID)) %>%
  # add back leading zeroes
  mutate(ID = str_pad(ID, width = 12, side = "left", pad = "0"))

# --- NEW (S3 or local via aws.s3) ---
# s3read_using downloads once to a local temp file; reading that temp file
# twice (header peek + selective read) costs one network round-trip, not two.
read_ejscreen_new <- function(f) {
  col_names_new <- names(readr::read_csv(f, n_max = 0, show_col_types = FALSE))
  keep_new <- col_names_new[grepl(VAR_NAME, col_names_new, ignore.case = TRUE)]
  readr::read_csv(
    f,
    col_select     = c("ID", "ST_ABBREV", all_of(keep_new)),
    col_types      = cols(ID = col_character(), .default = col_guess()),
    show_col_types = FALSE
  ) %>%
  # add back leading zeroes
  mutate(ID = str_pad(ID, width = 12, side = "left", pad = "0"))
}

ejscreen_new <- if (startsWith(path_new, "s3://")) {
  aws.s3::s3read_using(read_ejscreen_new, object = path_new)
} else {
  read_ejscreen_new(path_new)
}
```

```{r build-comparisons, include=FALSE}
# ---------------------------------------------------------------------------
# Build the core comparison frame and pre-compute difference columns.
# Frees both source datasets as soon as they're no longer needed.
# ---------------------------------------------------------------------------

# Case-insensitive match for the raw score column
col_old_name <- names(ejscreen_old)[tolower(names(ejscreen_old)) == tolower(VAR_NAME)]
col_new_name <- names(ejscreen_new)[tolower(names(ejscreen_new)) == tolower(VAR_NAME)]
col_old_name <- if (length(col_old_name)) col_old_name[1] else VAR_NAME
col_new_name <- if (length(col_new_name)) col_new_name[1] else VAR_NAME

COL_ORIG <- paste0(VAR_NAME, "_original")
COL_UPD  <- paste0(VAR_NAME, "_updated")

# join data frames based on ID column 
ejscreen_raw_comp <- dplyr::full_join(
  ejscreen_old %>% select(ID, all_of(col_old_name)) %>%
    rename(!!COL_ORIG := all_of(col_old_name)),
  ejscreen_new %>% select(ID, all_of(col_new_name)) %>%
    rename(!!COL_UPD  := all_of(col_new_name)),
  by = "ID"
)

# Attach state abbreviation (needed for region facets)
state_lookup <- bind_rows(
  ejscreen_old %>% select(ID, ST_ABBREV),
  ejscreen_new %>% select(ID, ST_ABBREV)
) %>% distinct(ID, .keep_all = TRUE)

ejscreen_raw_comp <- dplyr::left_join(ejscreen_raw_comp, 
                                      state_lookup, by = "ID")
rm(state_lookup)

# Slim pct_diff to just what the map section needs (ID + two difference cols)
# so the left-join with a 220k-row sf doesn't drag in a dozen extra columns.
pct_diff_map <- ejscreen_raw_comp %>%
  mutate(
    diff_new_minus_old = .data[[COL_UPD]] - .data[[COL_ORIG]],
    abs_diff           = abs(diff_new_minus_old),
    pct_diff_estimate  = 100 * (abs_diff /
      ((.data[[COL_UPD]] + .data[[COL_ORIG]]) / 2))
  ) %>%
  select(ID, pct_diff_estimate, abs_diff)
```

---


## Summary

```{r summary-blurb-calc, include=FALSE}
diff_vec   <- ejscreen_raw_comp[[COL_UPD]] - ejscreen_raw_comp[[COL_ORIG]]
pct_diff_vec <- 100 * abs(diff_vec) / ((ejscreen_raw_comp[[COL_UPD]] + ejscreen_raw_comp[[COL_ORIG]]) / 2)

n_total       <- nrow(ejscreen_raw_comp)
n_matched     <- sum(!is.na(diff_vec))

mean_orig     <- mean(ejscreen_raw_comp[[COL_ORIG]], na.rm = TRUE)
mean_upd      <- mean(ejscreen_raw_comp[[COL_UPD]],  na.rm = TRUE)
pct_change    <- 100 * (mean_upd - mean_orig) / mean_orig
direction     <- ifelse(mean_upd > mean_orig, "higher", "lower")

r_val         <- cor(ejscreen_raw_comp[[COL_ORIG]], ejscreen_raw_comp[[COL_UPD]],
                      use = "na.or.complete")
r2_val        <- r_val^2

mean_abs_diff <- mean(abs(diff_vec), na.rm = TRUE)
mean_abs_pct_dif <- mean(pct_diff_vec, na.rm = T)
q3_pct_diff   <- quantile(pct_diff_vec, probs = 0.75, na.rm = TRUE)

n_missing_orig_only <- sum(is.na(ejscreen_raw_comp[[COL_ORIG]]) & !is.na(ejscreen_raw_comp[[COL_UPD]]))
n_missing_upd_only  <- sum(!is.na(ejscreen_raw_comp[[COL_ORIG]]) & is.na(ejscreen_raw_comp[[COL_UPD]]))
n_missing_both       <- sum(is.na(ejscreen_raw_comp[[COL_ORIG]]) & is.na(ejscreen_raw_comp[[COL_UPD]]))
n_missing_total      <- n_missing_orig_only + n_missing_upd_only + n_missing_both
```

This scripted pulled the original EJScreen values from: `r path_old`; and new values from: `r path_new`.

Across `r format(n_total, big.mark = ",")` block groups (`r format(n_matched, big.mark = ",")` with values in both versions), updated `r VAR_NAME` scores are on average `r round(abs(pct_change), 2)`% `r direction` than the original values (mean absolute percent difference of `r round(mean_abs_pct_dif, 2)`; **Fig. 2**). The two versions remain tightly correlated (Pearson r² = `r round(r2_val, 2)`; **Fig. 1**), though the top quartile of block groups shows percent differences of `r round(q3_pct_diff, 2)`% or more (**Map 1**). Data availability also shifted slightly: `r format(n_missing_total, big.mark = ",")` block groups are missing a score in at least one version (`r n_missing_orig_only` missing only from the original, `r n_missing_upd_only` missing only from the updated, and `r n_missing_both` missing from both; **Map 2**).

# 1 · Score Comparisons

## Figure 1 – Relationship plot

```{r figure-1, fig.width=6, fig.height=6}
ggplot(ejscreen_raw_comp,
       aes(x = .data[[COL_ORIG]], y = .data[[COL_UPD]])) +
  geom_point(alpha = 0.4, size = 0.7, colour = "#526489") +
  stat_poly_eq(
    aes(label = paste(after_stat(eq.label), after_stat(rr.label),
                      sep = "*\", \"*")),
    formula   = y ~ x,
    parse     = TRUE,
    label.x   = "left",
    label.y   = "top"
  ) +
  geom_abline(intercept = 0, slope = 1, colour = "red", lty = "dashed") +
  theme_bw() +
  labs(x     = "Original EJScreen Value",
       y     = "Updated EJScreen Value",
       title = paste0("Relationship Plot: ", VAR_NAME))
```

This scatterplot compares each block group's original and updated `r VAR_NAME` 
value directly against one another (n = `r sum(!is.na(ejscreen_raw_comp[[COL_ORIG]]) & !is.na(ejscreen_raw_comp[[COL_UPD]]))` 
matched block groups). Points falling on the red dashed line indicate no 
change between versions; points above or below the line show the direction 
and magnitude of the shift.

*Note: the R² shown on the plot comes from OLS regression of updated ~ original
and equals Pearson r² only when both columns are on the same scale with no
offset. The red dashed line is the 1:1 reference, not the regression line.*

## Table 1 – Summary statistics
```{r table-1}
# compute stats
stats <- compute_stats(ejscreen_raw_comp, COL_ORIG, COL_UPD)

# extract the summary stats
summary_stats <- stats %>% 
  select(min_a:num_NA_var_b) %>%
  pivot_longer(everything(), 
               names_to = c(".value", "group"), 
               names_pattern = "(.*)_(a|b)$")%>%
  mutate(group = case_when(group == "a" ~ "Updated", 
                           group == "b" ~ "Original")) %>%
  rename(NA_obs= num_NA_var) %>%
  mutate(across(min:max, ~ round(., digits = 2)))

# format for knitting
summary_stats %>%
  knitr::kable(align = c("l", "r")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE)
```

This table summarizes the distribution of `r VAR_NAME` values in each dataset. 
The original data has `r summary_stats$NA_obs[summary_stats$group == "Original"]` 
missing observations, compared to `r summary_stats$NA_obs[summary_stats$group == "Updated"]` 
in the updated data.

## Table 2 – Comparison statistics
```{r}
# format the comparion stats
stats %>%
  select(!(min_a:num_NA_var_b)) %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
  mutate(
    Metric = recode(Metric,
                    matched_rows      = "Matched rows",
                    rows_used         = "Rows used",
                    dropped           = "Dropped (NA in either column)",
                    num_NA_vara       = "NAs in original dataset",
                    num_NA_varb       = "NAs in updated dataset",
                    mean_abs_diff     = "Mean absolute difference (upd − orig)",
                    median_abs_diff   = "Median absolute difference",
                    rmse              = "RMSE",
                    pearson_r         = "Pearson r",
                    r_sq              = "R²",
                    greatest_pos_diff = "Greatest positive difference (upd − orig)",
                    smallest_neg_diff = "Smallest negative difference (upd − orig)"
    ),
    Value = round(Value, 2)
  ) %>%
  knitr::kable(align = c("l", "r")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE)
```

This table quantifies agreement between versions. The two datasets show a 
Pearson r of `r round(stats$pearson_r, 2)` (r² = `r round(stats$r_sq, 2)`), 
with a mean absolute difference of `r round(stats$mean_abs_diff, 2)`.


## Figure 2 – Violin / boxplot

```{r figure-2, fig.width=7, fig.height=5}
ejscreen_raw_comp %>%
  select(ID, all_of(c(COL_ORIG, COL_UPD))) %>%
  pivot_longer(-ID, names_to = "version", values_to = "value") %>%
  mutate(version = recode(version,
    !!COL_ORIG := "Original",
    !!COL_UPD  := "Updated"
  )) %>%
  ggplot(aes(x = version, y = value, fill = version)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, alpha = 0.6, outlier.size = 0.5) +
  scale_fill_manual(values = cat_palette_pastel(2)) +
  labs(x     = "",
       y     = paste0(VAR_NAME, " value"),
       title = paste0("Distribution Comparison: ", VAR_NAME)) +
  epic_chart_theme +
  theme(legend.position = "none")
```

This plot compares the overall shape and spread of `r VAR_NAME` values, with 
means of `r round(stats$mean_b, 2)` (original) and `r round(stats$mean_a, 2)` 
(updated), highlighting any shifts in central tendency or variability.


```{r region-codes, include=FALSE}
# TODO - I'm not actually sure how useful this break down is. 
west_state_codes      <- c("04","08","16","30","32","35","49","56",
                           "02","06","15","41","53")
northeast_state_codes <- c("09","23","25","33","44","50","34","36","42")
midwest_state_codes   <- c("17","18","26","39","55","19","20","27",
                           "29","31","38","46")
south_state_codes     <- c("10","11","12","13","24","37","45","51","54",
                           "01","21","28","47","05","22","40","48")
all_conus_codes       <- c(west_state_codes, northeast_state_codes,
                           midwest_state_codes, south_state_codes)

assign_region <- function(fips2) {
  case_when(
    fips2 %in% west_state_codes      ~ "West",
    fips2 %in% northeast_state_codes ~ "Northeast",
    fips2 %in% midwest_state_codes   ~ "Midwest",
    fips2 %in% south_state_codes     ~ "South",
    TRUE                             ~ "Territory / Other"
  )
}
```

```{r figure-3, fig.width=11, fig.height=5}
# ejscreen_raw_comp %>%
#   mutate(region = assign_region(substr(ID, 1, 2))) %>%
#   pivot_longer(all_of(c(COL_ORIG, COL_UPD)),
#                names_to = "version", values_to = "value") %>%
#   mutate(version = recode(version,
#     !!COL_ORIG := "Original",
#     !!COL_UPD  := "Updated"
#   )) %>%
#   ggplot(aes(x = version, y = value, fill = version)) +
#   geom_violin(alpha = 0.5) +
#   geom_boxplot(width = 0.25, alpha = 0.6, outlier.size = 0.3) +
#   facet_wrap(~region, nrow = 1) +
#   scale_fill_manual(values = cat_palette_pastel(2)) +
#   labs(x     = "",
#        y     = paste0(VAR_NAME, " value"),
#        title = paste0("Regional Score Distributions: ", VAR_NAME)) +
#   epic_chart_theme +
#   theme(legend.position = "none",
#         strip.text      = element_text(face = "bold"))
```

---

# 2 · Pipeline Column Audit

```{r column-audit}
# removing extra strings from percentiles so we can actually 
# convert this column to numeric
ejscreen_old_audit <- ejscreen_old %>%
  mutate(across(starts_with("T_"), ~ sub(" %ile", "", .)))
ejscreen_new_audit <- ejscreen_new %>%
    mutate(across(starts_with("T_"), ~ sub(" %ile", "", .)))

# locating the columns to audit 
audit_cols_old <- names(ejscreen_old)[
  grepl(VAR_NAME, names(ejscreen_old), ignore.case = TRUE)
]

# creating audit table function 
audit_table <- purrr::map_dfr(audit_cols_old, function(col) {
  col_match <- names(ejscreen_new_audit)[tolower(names(ejscreen_new_audit)) == tolower(col)]

  if (!length(col_match)) {
    return(tibble(column = col, n_changed = NA_integer_,
                  mean_delta = NA_real_, status = "MISSING IN NEW"))
  }

  col_b <- col_match[1]
  merged_audit <- dplyr::inner_join(
    ejscreen_old_audit %>% select(ID, val_a = all_of(col)),
    ejscreen_new_audit %>% select(ID, val_b = all_of(col_b)),
    by = "ID"
  ) %>%
    # calculate deltas (b - a) and percent diffs
    mutate(val_a = suppressWarnings(as.numeric(val_a)),
           val_b = suppressWarnings(as.numeric(val_b)),
           delta = val_b - val_a, 
           pct_diff = 100*((delta)/((val_b + val_a)/2))) %>%
    # replace inf with NAs
    mutate(pct_diff = if_else(is.infinite(pct_diff), NA, pct_diff))
  
  # how many actually changed?
  n_changed <- merged_audit %>%
    filter(!(is.na(val_a) & is.na(val_b))) %>%
    filter(is.na(delta) | delta != 0) %>%
    nrow()

  tibble(column     = col,
         n_changed  = n_changed,
         mean_delta = mean(merged_audit$delta, na.rm = TRUE), 
         mean_pct_diff = mean(merged_audit$pct_diff, na.rm = TRUE))
})

# generate table, rename columns 
audit_table %>%
  arrange(desc(column)) %>%
  mutate(mean_delta = round(mean_delta, 6)) %>%
  rename(Column                   = column,
         `# Block Groups Changed` = n_changed,
         `Mean Delta (Upd − Orig)` = mean_delta, 
         `Mean Percent Diff (Upd − Orig)` = mean_pct_diff) %>%
  knitr::kable(align = c("l", "r", "r", "r")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE) %>%
  # removed the "OK" field 
  kableExtra::row_spec(which(audit_table$status != "OK"),
                       bold = TRUE, color = "white", background = "#c0392b")

# free up some memory
rm(ejscreen_old, ejscreen_new, ejscreen_old_audit, 
   ejscreen_new_audit, audit_cols_old)

invisible(gc())
```

This table audits every column in the original dataset containing `r VAR_NAME`, 
checking whether a matching column exists in the updated dataset and, where it 
does, how many block groups changed. Of the `r nrow(audit_table)` columns 
checked, `r sum(audit_table$status == "MISSING IN NEW", na.rm = TRUE)` 
could not be matched to the updated dataset and are flagged in *Map 2*. 

Mirrors compare_ejscreen.py run_audit()

---

# 3 · Difference Maps

```{r build-fips-xwalk-prep, include=FALSE}
fips_xwalk <- tigris::fips_codes %>%
  select(state, state_code) %>%
  distinct()

postal_all <- fips_xwalk %>%
  # these are minor outlying islands  - I don't think we would ever have 
  # data in EJScreen for these
  filter(state != "UM") %>%
  pull(state) %>% 
  unique()

# grabbing geometries 
bg_all <- postal_all %>%
  purrr::map(~tigris::block_groups(.x, cb = TRUE, progress_bar = FALSE)) %>%
  dplyr::bind_rows() %>%
  select(GEOID, STATEFP, geometry) %>%
  # shift_geometry before join so we transform the smaller pre-join frame
  # tigris::shift_geometry(geoid_column = "GEOID") %>%
  sf::st_transform(crs = "WGS84") %>%
  dplyr::left_join(pct_diff_map, by = c("GEOID" = "ID"))
```

## Map 1 - Largest Percent Differences (>= 3rd Quantile)
```{r map-percent-difference, fig.width=7, fig.height=5}
# finding 3rd quantile for the % difference estimates
q3 <- quantile(bg_all$pct_diff_estimate, probs = 0.75, na.rm = T)

# mappy time 
bg_over_q3 <- bg_all %>%
  filter(pct_diff_estimate >= q3) %>%
  mutate(pct_diff_estimate = round(pct_diff_estimate, 2))
mapview(bg_over_q3,
        zcol = "pct_diff_estimate", canvas = TRUE)@map
```
This map highlights the `r nrow(bg_over_q3)` block groups with the largest 
percent differences between original and updated `r VAR_NAME` scores — 
those at or above the 75th percentile (>= `r round(q3, 1)`%). Clusters on 
the map may point to regions where the underlying data or methodology 
changed most.

## Map 2 - Missing Scores 
```{r missing-scores-map, fig.width=7, fig.height=5}
# isolating geoids and geometries
bg_all_simp <- bg_all %>%
  select(GEOID) %>%
  st_as_sf() 

# we don't need this file anymore
rm(bg_all)

# create df of missing observations
missing_map <- ejscreen_raw_comp %>%
  rename(GEOID = ID) %>%
  # isolate for NAs in the original or updated columns 
  filter(is.na(.data[[paste0(VAR_NAME, "_original")]]) | is.na(.data[[paste0(VAR_NAME, "_updated")]])) %>%
  # identify what component is missing 
  mutate(`Missing Status` = case_when(is.na(.data[[paste0(VAR_NAME, "_original")]]) & is.na(.data[[paste0(VAR_NAME, "_updated")]]) ~ "Missing - Both", 
                                    is.na(.data[[paste0(VAR_NAME, "_original")]]) & !(is.na(.data[[paste0(VAR_NAME, "_updated")]])) ~ "Missing - Original Only",
                                    TRUE ~ "Missing - Updated Only")) %>%
  # bring in geometries
  left_join(., bg_all_simp, by = "GEOID") %>%
  st_as_sf() %>%
  st_cast(., "MULTIPOLYGON") %>%
  st_transform(., crs = 4326) %>%
  mutate(`Missing Status` = as.factor(`Missing Status`))


pal <- colorFactor(brewer.pal(3, "Set2"), domain = missing_map$`Missing Status`)

# an issue 
mapview(missing_map,
        zcol = "Missing Status", canvas = TRUE, 
        col.regions = pal(missing_map$`Missing Status`), 
        legend = T)

# ejscreen_raw_comp is no longer needed after this section
rm(ejscreen_raw_comp)
invisible(gc())
```

This map shows the `r nrow(missing_map)` block groups missing a `r VAR_NAME` 
score in the original data, the updated data, or both. Of these, 
`r sum(missing_map$missing_status == "missing from original")` are missing 
only from the original dataset, `r sum(missing_map$missing_status == "missing from updated")` 
only from the updated dataset, and `r sum(missing_map$missing_status == "both")` 
are missing from both.

---

# 4 · Works in Progress

## Figure 4 – Percentile scatter

```{r figure-4, fig.width=6, fig.height=6}
# t_pattern    <- paste0("^T_", VAR_NAME, "$")
# col_names_t  <- names(readr::read_csv(path_old, n_max = 0, show_col_types = FALSE))
# t_col_old    <- col_names_t[grepl(t_pattern, col_names_t, ignore.case = TRUE)]
# 
# if (length(t_col_old) == 1) {
#   t_old <- readr::read_csv(
#     path_old,
#     col_select     = c("ID", all_of(t_col_old)),
#     col_types      = cols(ID = col_character(), .default = col_character()),
#     show_col_types = FALSE
#   )
# 
#   read_t_new <- function(f) {
#     cn <- names(readr::read_csv(f, n_max = 0, show_col_types = FALSE))
#     t_col_new <- cn[grepl(t_pattern, cn, ignore.case = TRUE)]
#     if (!length(t_col_new)) return(NULL)
#     readr::read_csv(f, col_select = c("ID", all_of(t_col_new)),
#                     col_types = cols(ID = col_character(), .default = col_character()),
#                     show_col_types = FALSE)
#   }
# 
#   t_new <- if (startsWith(path_new, "s3://")) {
#     aws.s3::s3read_using(read_t_new, object = path_new)
#   } else {
#     read_t_new(path_new)
#   }
# 
#   if (!is.null(t_new)) {
#     t_percentiles <- dplyr::inner_join(t_old, t_new, by = "ID") %>%
#       rename(tile_str_old = 2, tile_str_new = 3) %>%
#       mutate(tile_old = as.numeric(str_extract(tile_str_old, "\\d+")),
#              tile_new = as.numeric(str_extract(tile_str_new, "\\d+")))
# 
#     ggplot(t_percentiles, aes(x = tile_old, y = tile_new)) +
#       geom_point(alpha = 0.4, size = 0.7, colour = "#526489") +
#       geom_abline(intercept = 0, slope = 1, colour = "red", lty = "dashed") +
#       theme_bw() +
#       labs(x     = paste0("Original T_", VAR_NAME, " tile"),
#            y     = paste0("Updated T_", VAR_NAME, " tile"),
#            title = paste0("T-Percentile Comparison: ", VAR_NAME))
# 
#     rm(t_old, t_new, t_percentiles)
#   } else {
#     message("T_", VAR_NAME, " not found in updated dataset — skipping Figure 4.")
#     rm(t_old)
#   }
# } else {
#   message("T_", VAR_NAME, " not found in original dataset — skipping Figure 4.")
# }
# gc()
```

## Map 3 - Largest Absolute Differences (>= 3rd Quantile)
```{r map-abs-difference}
# # finding 3rd quantile 
# q3 <- quantile(bg_all$abs_diff, probs = 0.75, na.rm = T)
# 
# # mappy time 
# bg_over_q3 <- bg_all %>%
#   filter(abs_diff >= q3)
# mapview(bg_over_q3,
#         zcol = "abs_diff", canvas = TRUE)@map
```
